---
title: "assignment_11"
format: html
editor: visual
---


## Required libraries

```{r}
library(dplyr)
library(meta)
```

###Reading the data
```{r}
m_data <- readxl::read_excel("data/metaanalysis_data.xlsx")
```

## Running meta analysis for boys play with female toys, using both fixed and random effect

```{r}
m.raw_boys_play_female <- metacont(n.e=N_boys,
                  mean.e=Mean_boys_play_female,
                  sd.e=SD_boys_play_female,
                  n.c=N_boys,
                  mean.c=Mean_boys_play_male,
                  sd.c=SD_boys_play_male,
                  data=m_data,
                  studlab=paste(Study),
                  comb.fixed = TRUE,
                  comb.random = TRUE,
)
```

### Plotting forest for it
```{r}
m.raw_boys_play_female %>% forest()
```
### Plotting funnel
```{r}
m.raw_boys_play_female %>% funnel()
```
### Finding does authers gender effect it
```{r}
m.raw_boys_play_female %>% metareg(`Study` + `Male authors`)
m.raw_boys_play_female %>% metareg(`Study` + `Female authors`)
```

## Running meta analysis for girls play with male toys, using both fixed and random effect
```{r}
m.raw_girsl_play_male <- metacont(n.e=N_girls,
                       mean.e=Mean_girls_play_male,
                       sd.e=SD_girls_play_male,
                       n.c=N_girls,
                       mean.c=Mean_girls_play_female,
                       sd.c=SD_girls_play_female,
                       data=m_data,
                       studlab=paste(Study),
                       comb.fixed = TRUE,
                       comb.random = TRUE,
)
```
### Plotting forest for it
```{r}
m.raw_girsl_play_male %>% forest()
```
### Plotting funnel
```{r}
m.raw_girsl_play_male %>% funnel()
```
### Finding does authers gender effect it
```{r}
m.raw_girsl_play_male %>% metareg(`Study` + `Male authors`)
m.raw_girsl_play_male %>% metareg(`Study` + `Female authors`)
```
