---
title: "(R)Markdown"
author: "Wojciech Hardy; Łukasz Nawaro"
date: "3/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# Three key elements

## The YAML header

At the very top. Stores author and document information, configuration, global options, inputs, styles, format specifications, etc. This document uses:

<pre class="code-block">
---
title: "Quarto"
author: "Wojciech Hardy"
date: "4/11/2023"
format: html
execute:
  echo: fenced
---
</pre>

## Markdown

The markup language, with easy and nice editing options.

## Code chunks

Where the magic happens - codes whose output becomes a part of the document.

Let's start with Markdown!


---

# Basic formatting

Some basic text formatting includes `*Italics*` or `_Italics_` (*Italics*) and `**Bold**` or `__Bold__` (**Bold**) text. 

`Space between lines to break a paragraph

Like so`

Space between lines to break a paragraph

Like so

`Or end a line with a double space to break  
without starting a new paragraph`

Or end a line with a double space to break  
without starting a new paragraph

Superscripts can be done like so: `R^2^` R^2^

Subscripts can be done like so: `H~2~O` H~2~O

---

# Headers

`# Header 1`

# Header 1

`## Header 2`

## Header 2

`### Header 3`

### Header 3

`#### Header 4`

#### Header 4

`##### Header 5`

##### Header 5

---

# Lists

## Ordered

`1. Item 1`

`2. Item 2`

`2. Item 3 # Note the error in numbering`

1. Item 1

2. Item 2

2. Item 3 `# It's fine here though`


## Unordered

`* Item`

`* Another item`

* Item

* Another item

## Subitems

1. Item 1
    + Item 2
    - Item 3

---

# Tables

| Day       | Hour  | Group  |
|:------    |:-----:|-------:|
| Wednesday | 9:45  | 1      |
| Thursday  | 16:45 | 2      |
| Thursday  | 18:30 | 3      |

---

# Equations

You can insert equations with the same syntax as in LaTeX. E.g. within a sentence `$ \sum (x + 1) $` $\sum (x + 1)$  or as standalone with double `$$` at start and finish

$$\int_{-\infty}^{\infty}1/(b-a) dx$$
---

# Code chunks customisation

([Partially taken from here](https://bookdown.org/yihui/rmarkdown-cookbook/hide-one.html))

## Chunk naming (for reference)

```{r chunk_name}
"You just put the name after the language declaration"
```


## Using variables as parameters

```{r}
typical_width <- 4
typical_height <- 4
```

```{r, fig.width=typical_width, fig.height=typical_height}
plot(cars)
```

## Using variables within the text `'r variable'`

We have previously set a typical width to `r typical_width` and the typical height to `r typical_height`.

## Keeping your report nice and clear

### Hide the source code with `echo=FALSE`

```{r, echo=FALSE}

1+1

```

### Hide messages (e.g. when loading stuff) with `message=FALSE`

```{r, message=FALSE}

message("You will not see the message.")

```

### Hide warnings with `warning=FALSE`

```{r, warning = FALSE}

1:2 + 1:3

```

### Hide plots with `fig.show='hide'`


```{r}

plot(cars)

```

```{r, fig.show='hide'}

plot(cars)

```

### Hide everything from the chunk with `include=FALSE`

```{r, include=FALSE}

cat("Some words")
3 + 4
# comment
plot(cars)

```

### Hide the results with `results='hide'`

```{r, results='hide'}

typical_height*typical_width
  
```

### Generate Markdown content with R code with `results='asis'`

```{r, results='asis'}

for (i in 1:10) {
  cat("- Item", i, "\n")
}
```

---

